---
title: "Darwin Core mapping"
subtitle: "For: Inventory of alien macroinvertebrates in Flanders, Belgium"
author:
- Lien Reyserhove
- Peter Desmet
- Dimitri Brosens
- Sanne Govaert
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes how we map the checklist data to Darwin Core. The source file for this document can be found [here](https://github.com/trias-project/alien-macroinvertebrates/blob/master/src/dwc_checklist.Rmd).

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r}
library(tidyverse)      # To do data science
library(magrittr)       # To use %<>% pipes
library(here)           # To find files
library(janitor)        # To clean input data
library(readxl)         # To read Excel files
```

# Read source data

The data is maintained in [this Google Spreadsheet](https://docs.google.com/spreadsheets/d/1LeXXbry2ArK2rngsmFjz_xErwE1KwQ8ujtvHNmTVA6E/edit#gid=510059691).

Read the relevant worksheet (published as csv):

```{r read_source_data}
raw_data <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTl8IEk2fProQorMu5xKQPdMXl3OQp-c0f6eBXitv0BiVFZ3JSJCde0PtbFXuETgguf6vK8b43FDX1C/pub?gid=510059691&single=true&output=csv", show_col_types = FALSE)
```

Copy the source data to the repository to keep track of changes:

```{r}
write_csv(raw_data, here("data", "raw", "alien_macroinvertebrates_dump.csv"), na = "")
```



# Preprocessing: tidy data and add taxon ID's

To link taxa with information in the extension(s), each taxon needs a unique and relatively stable `taxonID`. We have created one in the form of `dataset_shortname:taxon:hash`, where `hash` is unique code based on scientific name and kingdom. Once this is created, it is added to the source data. 

```{r}
input_data <-
  raw_data %>%
  remove_empty("rows") %>%
  clean_names() %>%
  mutate(
    taxon_id = paste(
      "alien-macroinvertebrates-checklist",
      "taxon",
      .data$taxon_id_hash,
      sep = ":"
    )
  )
```

## Use full references

The raw data contains (abbreviated) citations in `raw_reference` (e.g. `van Haaren and Soors 2009`). We want to use full reference (e.g. `van Haaren T, Soors J (2009) Sinelobus stanfordi (Richardson, 1901): a new crustacean invader in Europe. Aquatic Invasions 4: 703–711, https://doi.org/10.3391/ai.2009.4.4.20`), which are stored in a separate sources file. We will join the two here, so we can use this information in the core and/or extensions.

```{r}
sources <- read.table("../data/raw/sources.tsv", sep = "\t", quote = "", colClasses = "character",  fileEncoding = "UTF8", header = T)
```

Clean `raw_reference` somewhat:

```{r}
input_data %<>% mutate(reference = recode(source,
  "Adam  and Leloup 1934" = "Adam and Leloup 1934", # Remove whitespace
  "Van  Haaren and Soors 2009" = "van Haaren and Soors 2009", # Remove whitespace and lowercase "van"
  "This study" = "Boets et al. 2016",
  "Nyst 1835; Adam 1947" = "Nyst 1835 | Adam 1947" )
)
```

Left join `input_data` with `sources` on `references` = `citation`:

```{r}
input_data %<>% 
  left_join(sources, by = c("source" = "citation"))
```

Show result (note: `Boets. et al. unpub data` and `Collection RBINS` don't have a full reference):

```{r}
input_data %>% 
  mutate(full_reference = substr(full_reference, 1, 50)) %>% # Shorten full_reference to make it easier to display
  select(reference, full_reference) %>%
  group_by_all %>%
  summarize(records = n()) %>%
  arrange(reference)
```




# Darwin Core mapping

## Taxon core

Create a dataframe with unique taxa only (ignoring multiple distribution rows). Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml).

```{r}
taxon <-
  input_data %>%
  distinct(taxon_id, .keep_all = TRUE) %>%
  mutate(
    language = "en",
    license = "http://creativecommons.org/publicdomain/zero/1.0/",
    rightsHolder = "Ghent University Aquatic Ecology",
    #accessRights = "https://www.inbo.be/en/norms-data-use",
    datasetID = "https://doi.org/10.15468/yxcq07",
    #institutionCode = "",
    datasetName = "Inventory of alien macroinvertebrates in Flanders, Belgium",
    taxonID = taxon_id,
    scientificName = scientific_name,
    kingdom = kingdom,
    phylum = phylum,
    #class = class,
    order = order,
    family = family,
    genus = genus,
    taxonRank = taxon_rank,
    nomenclaturalCode = nomenclatural_code,
    .keep = "none"
  ) %>% 
  arrange(taxonID) %>% 
  select(
    "language", "license", "rightsHolder", # "accessRights", 
    "datasetID", #"institutionCode",
    "datasetName", "taxonID", "scientificName", "kingdom", 
    "phylum", # "class", 
    "order", "family", "genus", "taxonRank", "nomenclaturalCode"
  )
```

## Distribution extension

Create a dataframe with all data (including multiple distributions). Map the data to [Species Distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml).

Information for `eventDate` is contained in `date_first_observation` and `date_last_observation`, which we will express here in an ISO 8601 date format `yyyy/yyyy` (`start_date/end_date`).

Not all cells for `date_first_observation` (DFO) and/or `date_last_observation` (DLO) are populated. So, we used the following rules for those records: 

***case 1.*** If `DFO` is empty and `DLO` is empty, `eventDate` is `NA` 
***case 2.***  If `DFO` is empty and `DLO` is not empty: eventDate = `/DLO`
***case 3.*** If `DFO` is not empty and `DLO` is empty, eventDate is `DFO/`

```{r}
distribution <-
  input_data %>%
  # pathway
  pivot_longer(
    names_to = "key",
    values_to = "pathway",
    starts_with("introduction_pathway"),
    values_drop_na = FALSE) %>%
  filter( # keep NA value for species with no pathway provided
    !is.na(pathway) |
      (is.na(pathway) & key == "introduction_pathway_1")
    ) %>%
  # other terms
  mutate(
    taxonID = taxon_id,
    locationID = case_when(
      location == "Flanders" ~ "ISO_3166-2:BE-VLG",
      location == "Wallonia" ~ "ISO_3166-2:BE-WAL",
      location == "Brussels" ~ "ISO_3166-2:BE-BRU"
    ),
    locality = case_when(
      location == "Flanders" ~ "Flemish Region",
      location == "Wallonia" ~ "Walloon Region",
      location == "Brussels" ~ "Brussels-Capital Region"
    ),
    countryCode = country_code,
    occurrenceStatus = occurrence_status,
    establishmentMeans = establishment_means,
    degreeOfEstablishment = degree_of_establishment,
    eventDate = case_when(
      is.na(date_first_observation) & is.na(date_last_observation) ~ NA,
      is.na(date_first_observation) ~ paste0("/", date_last_observation),
      is.na(date_last_observation) ~ paste0(date_first_observation, "/"),
      !is.na(date_first_observation) & !is.na(date_last_observation) ~ 
        paste(date_first_observation, date_last_observation, sep = "/")
    ),
    source = full_reference,
    occurrenceRemarks = occurrence_remarks
  ) %>%
  select(
    "taxonID", "locationID", "locality", "countryCode", "occurrenceStatus",
    "establishmentMeans", "degreeOfEstablishment", "pathway", 
    "eventDate", "source", "occurrenceRemarks"
  ) %>%
  arrange(taxonID)
```


## Species profile extension

In this extension we will express broad habitat characteristics of the species (e.g. `isTerrestrial`).

Create a dataframe with unique taxa only (ignoring multiple distribution rows).
Only keep records for which `terrestrial`, `marine` and `freshwater` is not empty.

Map the data to [Species Profile](http://rs.gbif.org/extension/gbif/1.0/speciesprofile.xml).

```{r}
species_profile <-
  input_data %>%
  distinct(taxon_id, .keep_all = TRUE) %>% 
  filter(
    !is.na(terrestrial) |
      !is.na(marine) |
      !is.na(freshwater)
  ) %>% 
  mutate(
    .keep = "none",
    taxonID = taxon_id,
    isMarine = marine,
    isFreshwater = freshwater,
    isTerrestrial = terrestrial
  ) %>% 
  arrange(taxonID)
```

## Description extension

In the description extension we want to include the native range of a species

```{r}
description <-
  input_data %>% 
  # unique taxa only (ignoring multiple distribution rows)
  distinct(taxon_id, .keep_all = TRUE) %>% 
  # Separate values on `|` 
  mutate(native_range = strsplit(native_range, "\\|")) %>% 
  unnest(native_range) %>% 
  filter(!is.na(native_range)) %>% 
  mutate(
    .keep = "none",
    taxonID = taxon_id,
    description = str_trim(native_range),
    type = "native range",
    source = full_reference,
    language = "en"
    ) %>% 
  select("taxonID", "description", "typ", "source", "language") %>% 
  arrange(taxonID)
```

# Save to CSV:

```{r}
write_csv(taxon, "../data/processed/dwc_checklist/taxon.csv", na = "")
write_csv(distribution, "../data/processed/dwc_checklist/distribution.csv", na = "")
write_csv(species_profile, "../data/processed/dwc_checklist/speciesprofile.csv", na = "")
write_csv(description, "../data/processed/dwc_checklist/description.csv", na = "")
```

